\cleardoublepage
\phantomsection
\addcontentsline{toc}{chapter}{Acknowledgements}
\begin{acknowledgements}
    \begin{spacing}{1.25}
    First off I'd like to thank my wonderful supervisors, David and Helen, for all their guidance and support on my DPhil journey. No matter how busy they've been, they've always found time for me. I couldn't have asked for a better team than them.

    I'd also like to thank Elizabeth Yankovsky and three anonymous reviewers for their invaluable feedback on the contents of chapters two, three \& four.
    
    For introducing me to the world of OSNAP and her help with chapter five, I'd like to say a huge thanks to Isabela Le Bras.
    
    For taking me to sea and successfully getting me back home when the plague took hold, thanks to the RAPID AMOC team and the crew on board JC192.
    
    Starting out, I hadn't realised the definition of symmetric instability was so contentious, so thanks to Brian Hoskins and James Maddison for reassuring me it really is as simple as $fQ < 0$.
    
    Behind the scenes, I've been supported by some wonderful friends, so in no particular order, I'd like to say a huge thank you to Beth, Nick, Stuart and anyone else I've spent too much time in the AOPP kitchen with, the Oxford Ocean Climate group, Nilani, the Breakfast Club, Simon and the Oxford University Yacht Club, the 2018 cohort of the NERC DTP, and Lilly.
    
    Finally, a heartfelt thanks to my family who've been my biggest supporters throughout.

    I am grateful for the financial support of the Natural Environment Research Council (NE/L002612/1). Also, this work used the ARCHER2 UK National Supercomputing Service (\url{https://www.archer2.ac.uk}).
    \end{spacing}
\end{acknowledgements}